---
output:
  github_document:
    html_preview: false
always_allow_html: true
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
pkgload::load_all(".", quiet = TRUE)

device <- Sys.getenv(
  "RLIMITE_README_DEVICE",
  if (torch::cuda_is_available()) "cuda" else "cpu"
)

# Reasoning traces are long by design; the landing page shows the opening.
show_trace <- function(text, max_lines = 16L) {
  lines <- strsplit(text, "\n", fixed = TRUE)[[1]]
  if (length(lines) > max_lines) {
    lines <- c(
      lines[seq_len(max_lines)],
      sprintf("[... %d further lines]", length(lines) - max_lines)
    )
  }
  cat(lines, sep = "\n")
}
```

# Rlimite: a math model on the machine you already have

<!-- badges: start -->
[![R-CMD-check](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/R-CMD-check.yaml)
[![Real model tests](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/real-model.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/real-model.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rlimite)](https://sounkou-bioinfo.r-universe.dev/Rlimite)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

**One package, one argument: a trained model is a file, and a file is enough.
Solving a mathematical problem should not require a Python interpreter, a
serving process, or a second copy of the weights in another runtime.**

Rlimite implements the
[Limite](https://github.com/paradigma-inc/limite-violetto) decoder architecture
directly in R with [`torch`](https://torch.mlverse.org/). The published
safetensors and tokenizer are read where they already live, in the Hugging Face
cache, and executed in the calling R session. There is no `reticulate` bridge,
no vLLM, no model server, and no network round-trip per prompt. After the first
download the machine in front of you is the whole system, on CPU or on GPU.

Limite 1B - Violetto is math-specialized and lightly instruction-tuned for
independent single-turn problems. It is a solver, not a general chat assistant,
and this README does not pretend otherwise. Base and Base Soup are earlier
checkpoints of the same published family and are shown here doing what they
actually do.

## Installation

```r
install.packages(
  "Rlimite",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

Rlimite needs `{torch}` with its LibTorch runtime, plus `{safetensors}`,
`{hfhub}` and `{tok}`. Each published BF16 checkpoint is about 2.08 GB on disk,
before activations and caches.

## A file and a tokenizer are enough

Every example below is evaluated when this README is rendered, against the
pinned revisions in `limite_checkpoints()`. Nothing on this page is a
transcript written by hand.

```{r quickstart, cache = TRUE, cache.path = "README_cache/violetto/", cache.extra = list(device, subset(limite_checkpoints(), checkpoint == "Violetto")$revision)}
library(Rlimite)

violetto <- subset(limite_checkpoints(), checkpoint == "Violetto")
model <- limite_from_pretrained(
  violetto$model_id,
  revision = violetto$revision,
  device = device
)
tokenizer <- limite_tokenizer(violetto$model_id, revision = violetto$revision)

messages <- list(list(
  role = "user",
  content = "Compute 17 * 23. Show your reasoning."
))
solution <- limite_chat(
  model,
  messages,
  tokenizer = tokenizer,
  max_new_tokens = 256,
  temperature = 0
)
```

The model reasons before it commits to an answer, so the trace is long. Its
opening is the interesting part:

```{r quickstart-trace}
show_trace(solution)
```

Weights attach on the meta device and are then bound to the checkpoint tensors,
so loading never materializes a second initialized copy of the 1B parameters.
The safetensors mapping stays authoritative.

## The device is an argument, not an architecture

The same call runs on either processor. `device = "cuda"` moves the parameters
onto the GPU, and generation follows the weights rather than being rewritten
around them:

```r
model <- limite_from_pretrained(
  "paradigma-inc/limite-1b-violetto",
  device = "cuda"
)
```

`limite_generate()` reads its device from the embedding weights, so no calling
code carries a device argument at all. This page reports whichever processor
rendered it, and `RLIMITE_README_DEVICE` forces the choice:

```{r throughput}
prompt <- limite_chat_prompt(messages)
prompt_ids <- tokenizer$encode(prompt, add_special_tokens = FALSE)$ids

elapsed <- system.time(
  tokens <- limite_generate(
    model,
    prompt_ids,
    max_new_tokens = 32L,
    temperature = 0,
    eos_token_id = NULL
  )
)[["elapsed"]]

data.frame(
  device = device,
  prompt_tokens = length(prompt_ids),
  new_tokens = length(tokens),
  seconds = round(elapsed, 1),
  tokens_per_second = round(length(tokens) / elapsed, 2)
)
```

This is a single warm greedy decode with a key/value cache, not a kernel
microbenchmark, and it should be read as such.

## Watching it decode

Generation is not a black box that returns after a minute. `limite_generate()`
takes a `callback` fired on every sampled token, so a solution can stream into
the console while it is produced:

```{r streaming}
invisible(limite_generate(
  model,
  prompt_ids,
  max_new_tokens = 24L,
  temperature = 0,
  eos_token_id = NULL,
  callback = function(id) cat(tokenizer$decode(id, skip_special_tokens = TRUE))
))
```

## Nothing has to leave the machine

Once a revision is in the Hugging Face cache, `local_files_only = TRUE` removes
the network from the loading path entirely:

```r
model <- limite_from_pretrained(
  "paradigma-inc/limite-1b-violetto",
  local_files_only = TRUE
)
```

The same argument accepts a checkpoint directory or a single `.safetensors`
file, so an air-gapped machine loads weights copied in by hand, and a machine
with no Python installation at all is not a special case.

## The family is three checkpoints, not one

All three pinned revisions are loaded and given the same problem, with twelve
greedy tokens each. The comparison is the point: the base checkpoints continue
text, and only Violetto has been tuned to solve.

```{r checkpoints, cache = TRUE, cache.path = "README_cache/family/", cache.extra = list(device, limite_checkpoints()$revision)}
checkpoints <- limite_checkpoints()
checkpoints$opening <- NA_character_

for (i in seq_len(nrow(checkpoints))) {
  is_violetto <- identical(checkpoints$checkpoint[[i]], "Violetto")

  if (is_violetto) {
    current_model <- model
    current_tokenizer <- tokenizer
  } else {
    current_model <- limite_from_pretrained(
      checkpoints$model_id[[i]],
      revision = checkpoints$revision[[i]],
      device = device
    )
    current_tokenizer <- limite_tokenizer(
      checkpoints$model_id[[i]],
      revision = checkpoints$revision[[i]]
    )
  }

  ids <- limite_generate(
    current_model,
    prompt_ids,
    max_new_tokens = 12L,
    temperature = 0,
    eos_token_id = NULL
  )
  checkpoints$opening[[i]] <- current_tokenizer$decode(
    ids,
    skip_special_tokens = TRUE
  )

  if (!is_violetto) {
    rm(current_model, current_tokenizer)
    gc()
  }
}
```

```{r checkpoints-table, echo = FALSE}
knitr::kable(checkpoints[c("checkpoint", "profile", "revision", "opening")])
```

The revisions are immutable commit hashes rather than `main`, so this table and
the real-model CI matrix describe the same three artifacts.

## What the architecture actually required

The shared checkpoint contract is 48 decoder layers, grouped-query attention,
mixed global and sliding-window layers, and a 131,072-token context.

```{r architecture, echo = FALSE}
config <- limite_config()
unlist(config[c(
  "vocab_size",
  "hidden_size",
  "num_hidden_layers",
  "num_attention_heads",
  "num_key_value_heads",
  "max_position_embeddings"
)])
```

None of the following is optional for numerical agreement with the published
checkpoints, and each is implemented in R rather than deferred to a kernel
library:

- partial interleaved RoPE, with NoPE on the global layers
- value embeddings and per-head attention gates
- cross-stream attention
- MUDD history mixing with learned taps
- learned residual coefficients, and projection-scale folding at load time
- sigmoid-softcapped logits
- bounded key/value caches on the sliding-window layers

Public token IDs stay zero-based, matching the Hugging Face tokenizers. The
conversion to R torch's one-based embedding indices happens only at the model
boundary, so IDs handed to `limite_generate()` and read back from the
`token_ids` attribute line up with the Hub tokenizer exactly.

## What is not measured yet

The throughput table above reports one device: the one that rendered the page.
A CPU-versus-CUDA comparison on the same prompt, the same revision and the same
decode length is the honest next measurement, and it is not in this README
because it has not been run. Attention is a dense materialized product here
rather than a fused kernel, and no claim is made about how this stack compares
with a tuned serving runtime on the same weights. Those numbers belong here
once they exist.

## Model weights

Rlimite does not redistribute model weights. Access to and use of the
checkpoints remain subject to the terms attached to the selected model
repository.
