---
output:
  github_document:
    html_preview: false
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
pkgload::load_all(".", quiet = TRUE)
```

# Rlimite

[![R-CMD-check](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/R-CMD-check.yaml)
[![Real model tests](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/real-model.yaml/badge.svg)](https://github.com/sounkou-bioinfo/Rlimite/actions/workflows/real-model.yaml)
[![R-universe](https://sounkou-bioinfo.r-universe.dev/badges/Rlimite)](https://sounkou-bioinfo.r-universe.dev/Rlimite)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)

Rlimite implements the
[Limite](https://github.com/paradigma-inc/limite-violetto) decoder architecture
directly in R with [`torch`](https://torch.mlverse.org/). It loads the official
safetensors and tokenizer assets without Python, vLLM, or a model-serving
process.

Limite 1B - Violetto is a math-specialized, lightly instruction-tuned model for
independent single-turn problems. It is not intended to behave like a general
chat assistant. The Base and Base Soup checkpoints expose earlier members of
the same published model family.

## Published checkpoints

The README pins the exact revisions used for its evaluated examples and for the
real-model CI matrix.

```{r checkpoints}
checkpoints <- limite_checkpoints()
knitr::kable(checkpoints)
```

## Architecture

The shared checkpoint contract contains 48 decoder layers, grouped-query
attention, mixed global and sliding-window layers, and a 131,072-token context
limit.

```{r architecture}
config <- limite_config()
unlist(config[c(
  "vocab_size",
  "hidden_size",
  "num_hidden_layers",
  "num_attention_heads",
  "num_key_value_heads",
  "max_position_embeddings"
)])
```

The implementation includes partial interleaved RoPE, NoPE global layers,
value embeddings, per-head attention gates, cross-stream attention, MUDD
history mixing, learned residual coefficients, projection-scale folding,
sigmoid-softcapped logits, and bounded local-layer key/value caches.

## Installation

Install Rlimite from R-universe with:

```r
install.packages(
  "Rlimite",
  repos = c(
    "https://sounkou-bioinfo.r-universe.dev",
    "https://cloud.r-project.org"
  )
)
```

Rlimite requires `{torch}` with its LibTorch runtime, `{safetensors}`, `{hfhub}`,
and `{tok}`. A CUDA-enabled LibTorch installation is recommended, but all
examples below are evaluated on CPU. Each published BF16 checkpoint is about
2.08 GB before runtime activations and caches.

## Verify all three checkpoints

This chunk loads every pinned checkpoint and begins the same mathematical
problem with eight greedy tokens. It reports runtime-contract checks rather
than comparing the quality of differently trained checkpoints.

```{r model-matrix}
problem <- "Compute 17 * 23. Show your reasoning."
messages <- list(list(role = "user", content = problem))
checks <- data.frame(
  checkpoint = checkpoints$checkpoint,
  generated_tokens = integer(nrow(checkpoints)),
  valid_token_ids = logical(nrow(checkpoints))
)

for (i in seq_len(nrow(checkpoints))) {
  model <- limite_from_pretrained(
    checkpoints$model_id[[i]],
    revision = checkpoints$revision[[i]],
    device = "cpu"
  )
  tokenizer <- limite_tokenizer(
    checkpoints$model_id[[i]],
    revision = checkpoints$revision[[i]]
  )
  continuation <- limite_chat(
    model,
    messages,
    tokenizer = tokenizer,
    max_new_tokens = 8,
    temperature = 0,
    eos_token_id = NULL
  )
  token_ids <- attr(continuation, "token_ids")
  checks$generated_tokens[[i]] <- length(token_ids)
  checks$valid_token_ids[[i]] <- all(
    token_ids >= 0L & token_ids < model$config$vocab_size
  )

  if (i < nrow(checkpoints)) {
    rm(model, tokenizer)
    gc()
  }
}

stopifnot(all(checks$generated_tokens == 8L), all(checks$valid_token_ids))
knitr::kable(checks)
```

## Solve a problem with Violetto

The final model retained by the preceding evaluated chunk is Violetto. Its
published message template wraps the direct single-turn problem. Violetto emits
a long reasoning trace; the evaluated example extracts the boxed result from
the generated token stream so the landing page stays concise.

```{r violetto}
solution <- limite_chat(
  model,
  messages,
  tokenizer = tokenizer,
  max_new_tokens = 256,
  temperature = 0
)
boxed_answer <- regmatches(
  solution,
  regexpr("\\\\boxed\\{[^}]+\\}", solution, perl = TRUE)
)
stopifnot(length(boxed_answer) == 1L)
boxed_answer
```

## Token IDs and model assets

Public token IDs are zero-based, matching Hugging Face tokenizers. Rlimite
converts them to R torch's one-based embedding indices only at the model
boundary.

Rlimite does not redistribute model weights. Weight access and use remain
subject to the terms attached to the selected model repository.
